What We Eval in the Shadows

### 2021-04-16 paper/main.tex
Grammarly's proofreading of Section 4 (Usage metrics)

There are \PkgEvalCallSites \eval sites in \PkgPackages packages. The proportion
of packages calling \eval is \PkgPackagesRatio. Over half of these packages
have fewer than 3 sites, and with the exception \MaxEvalCallSitesPackage which
has \MaxEvalCallSitesCount sites, all packages contain fewer than
\MaxEvalCallSitesRest sites (\cf Fig~\ref{fig:pkgs-eval-callsites-hist} which
shows a histogram of sites per package). These sites appear in \PkgFunsWithEval
functions (\CranFunsWithEvalRatio of all functions in CRAN).

@@

For dynamic analysis, we run \CranRunnableScripts programs extracted from
\CranPackages packages. Any run that does not exercise \eval is discarded. This
left \packageNbruns runs from \packageCorpus packages. There were
\packageAllcalls calls in \packageTriggeredpkgs packages originating from
\PkgHitEvalCallSites unique sites. In terms of coverage, the data exercised
\PkgHitEvalCallSitesAvgRatio of sites, a coverage similar to the code coverage
metric for the packages which is \PkgCodeCoverage. The fact that not all sites
are exercised can be chalked down to incomplete tests and occasional analysis
failures (\PkgFailedProgramsRatio of programs crashed or timed out).
Fig~\ref{fig:traced-eval-callsites} shows the number of sites that were
exercised; coverage is unequal. Figure~\ref{freq} summarizes the frequency of
dynamic calls to \eval with the left column being the number of calls and the
right, the number of packages in that range. There are \packageFewcalls packages
with low \eval frequency, fewer than 100 calls; and a smaller, but still
significant, number of packages, \packageManycalls to be precise, that use \eval
more than 1,000 times. Package \packageMaxcallspack makes \packageMaxcalls calls
and thus accounts for over half of the observations.
Figure~\ref{tab:variantseval} summarizes the use of variants of \eval. For each
of the four variants, the first rows shows the number of sites in the corpus
(\emph{static}), the second row is the number of sites that were encountered
during analysis (\emph{exercised}), the last row is the number of calls
(\emph{invocations}). The overwhelming majority of sites and calls, are to \eval
itself, \c{eval.parent} is rare, and both \c{evalq} and \c{local} are barely
used at all. The difference between sites and exercised sites underscores the
limitations due to code coverage.

@@

Figure~\ref{cn}(a) shows normalized call counts per site; on the left are
average number of calls from a given site and given run, on the right are counts
of sites that fall in that range. For instance, \packageRunbinh sites are
invoked 2,000+ times per run. Larger numbers suggest loops or recursive contexts
-- this seems to be the exception as most \evals, \packageRunbina, are exercised
50 times or less. Figure~\ref{cn}(b) zooms in on low frequency sites. The x-axis
shows normalized calls and the y-axis is the number of sites for that value. Most
low-frequency \evals are invoked only once, about half as many are invoked
twice; after that the frequency quickly drops.

\Eval takes any value, but if its argument is not an expression, \eval returns
it unchanged. Expressions account for \packageCodepercent of arguments in our
corpus. More specifically, \packageSymbolpercent are \texttt{symbol}s (single
variables such as \c{x}), \packageLanguagepercent are \c{language} objects
(function calls such as \c{f(x)} or \c{x+1}), and \packageExpressionpercent are
\c{expression} objects (lists of expressions such as function calls or
symbols). Further inspection reveals that most symbols,
\packageGgplotsymbolpercent to be exact, come from a single site in the
\c{ggplot2} package and have the value \c{\_inherit}.\footnote{This models
inheritance in \c{ggproto}, one of the many object-oriented systems in
R that is used exclusively in the \c{ggplot2} graphics library.}

To estimate how much executable code is injected through \eval, we measure the number of nodes in the expressions; for example, \c{x+1} counts as 3. We
also measured string lengths of unparsed expressions, but these measurements
were dominated by the size of data objects which could range in the MBs. The
median argument size is \packageMedianszeval (due to the symbols) and the
average is \packageAvgszeval nodes. The largest \eval input observed is
\packageMaxszeval\,-- a significant chunk of code.
Figure~\ref{fig:sizedistribution} shows the distribution of sizes for arguments
of fewer or equal to 25 nodes. The x-axis is the size of arguments in number of
nodes, and the y-axis is the count of arguments with that size. The size drops
rapidly, with few observations larger than 15 nodes. The long tail is omitted
for legibility.

To estimate the work performed in \evals, we count instructions executed by the
interpreter. Most invocations perform relatively little work with
\packageSmalleventspct of \evals executing 50 or fewer instructions. The violin
plot of Figure~\ref{ev}(a) corresponds to \evals executing $\leq$ 50
instructions, it is dominated by trivial symbol look ups. Figure~\ref{ev}(b) has
the work-intensive \evals which go all the way to \packageMaxeventsRnd
instructions.

@@

We recorded \baseAllcalls \eval calls in the \BasePackages base libraries. This
comes from running 10\% randomly selected programs from the
\CranRunnableScriptsRnd extracted programs from CRAN packages, covering
\baseTriggeredevalpct of the \BaseEvalCallSites sites. Most of the calls
(\baseEvalsratio) are to the \c{eval} function and most arguments
(\baseCodepercent) are expressions. However, unlike CRAN, the majority
(\baseLanguagepercent) of the arguments are language objects. The median
argument size is \baseMedianszeval, which is more than for CRAN and the maximum
size is \baseMaxszeval, less than for CRAN. Small instruction counts ($<50$),
amount for \baseSmalleventspct of calls. A single site in the \c{match.arg}
function is responsible for \baseTopFuncPercent of all recorded calls. This
function provides a convenient way for argument verification using partial
matching. It is heavily relied on in R functions that use string arguments to
parameterize function behavior, a popular API mechanism in R. For example, in a
body of \c{center <- function(x, type = c("mean", "median", "trimmed"))} one
can get the value of \c{type} parameter using \c{type <- match.arg(type)}. A
call \c{center(x, "trim")} will match \c{type} to \c{"trimmed"}.

The challenge dealing with Base is that every program uses it. It can thus
generate extreme amounts of data but the data is quite predictable. For the
rest of the paper, we only mention Base when there are surprising
observations.

@@

In total, \kaggleAllcalls \eval were recorded, all to \eval function. Out of
\kaggleStaticeval sites from \KaggleWithEvals, only \kaggleTriggeredeval sites
in \kaggleNbruns scripts were hit. \KaggleFailedScripts scripts failed to run
and the other did not exercised the \eval sites. This is partially expected
as Kaggle code does not need to abide to any checks. Upon a manual inspection,
we observed that indeed the failing scripts were of a poor quality, often not
finished, using misspelled package names, hard-coded file paths or accessing
missing files.

Most call sites are invoked only once. Only one site is called \kaggleMaxcalls times.
Expressions amount for \kaggleCodepercent of arguments. Unlike CRAN corpus,
there are very few symbols (\kaggleSymbolpercent). Most expressions result from
calls to \c{parse}, thus most \evals start with strings. The median argument
size is \kaggleMedianszeval, which makes sense, as few arguments are only
symbols. The largest argument is \kaggleMaxszeval. The distribution of
instructions per eval is similar to CRAN; \kaggleSmalleventspct of \evals
execute fewer than 50 instructions. Manual inspection of \eval usage in Kaggle
suggest that it is consistent with the data obtained for CRAN. We do not
discuss it further.

@@

\subsection{Discussion}
These results show that \eval is central to the language implementation as it is
omnipresent in the relatively small Base library. The Base library is, in turn,
used by every single package. To avoid having Base dominate the result we have
excluded calls occurring there from the data we report when analyzing packages.
CRAN Packages, which are typically developed by experienced programmers, make
regular and varied use of \eval. They represent our most interesting data set as
these packages are the result of over 20 years of contributions by thousands of
authors. The code in packages is well maintained and relatively well tested.
Finally, Kaggle scripts are often written by less sophisticated users, and
likely perform simpler tasks, and thus have lesser need for \eval, and its uses
originate from strings as these are easier to manipulate for end-users. Due to
the relative paucity of data in Kaggle, we do not pursue the data set further.

>>>>

There are \PkgEvalCallSites \eval sites in \PkgPackages packages. The proportion
of packages calling \eval is \PkgPackagesRatio. Over half of these packages
have fewer than 3 sites, and with the exception \MaxEvalCallSitesPackage, which
has \MaxEvalCallSitesCount sites, all packages contain fewer than
\MaxEvalCallSitesRest sites (\cf Fig~\ref{fig:pkgs-eval-callsites-hist}, which
shows a histogram of sites per package). These sites appear in \PkgFunsWithEval
functions (\CranFunsWithEvalRatio of all functions in CRAN).

@@

For dynamic analysis, we run \CranRunnableScripts programs extracted from
\CranPackages packages. Any run that does not exercise \eval is discarded. This
left \packageNbruns runs from \packageCorpus packages. There were
\packageAllcalls calls in \packageTriggeredpkgs packages originating from
\PkgHitEvalCallSites unique sites. In terms of coverage, the data exercised
\PkgHitEvalCallSitesAvgRatio of sites, a coverage similar to the code coverage
metric for the packages, which is \PkgCodeCoverage. The fact that not all sites
are exercised can be chalked down to incomplete tests and occasional analysis
failures (\PkgFailedProgramsRatio of programs crashed or timed out).
Fig~\ref{fig:traced-eval-callsites} shows the number of sites that were
exercised; coverage is unequal. Figure~\ref{freq} summarizes the frequency of
dynamic calls to \eval, with the left column being the number of calls and the
right, the number of packages in that range. There are \packageFewcalls packages
with low \eval frequency, fewer than 100 calls, and a smaller, but still
significant, number of packages, \packageManycalls to be precise, that use \eval
more than 1,000 times. Package \packageMaxcallspack makes \packageMaxcalls calls
and thus accounts for over half of the observations.
Figure~\ref{tab:variantseval} summarizes the use of variants of \eval. For each
of the four variants, the first row shows the number of sites in the corpus
(\emph{static}), the second row is the number of sites that were encountered
during analysis (\emph{exercised}), the last row is the number of calls
(\emph{invocations}). The overwhelming majority of sites and calls are to \eval
itself, \c{eval.parent} is rare, and both \c{evalq} and \c{local} are barely
used at all. The difference between sites and exercised sites underscores the
limitations due to code coverage.

@@

Figure~\ref{cn}(a) shows normalized call counts per site; on the left are
the average number of calls from a given site and given run, on the right are the counts of sites that fall in that range. For instance, \packageRunbinh sites are
invoked 2,000+ times per run. Larger numbers suggest loops or recursive contexts
-- this seems to be the exception as most \evals, \packageRunbina, are exercised
50 times or less. Figure~\ref{cn}(b) zooms in on low frequency sites. The x-axis
shows normalized calls, and the y-axis is the number of sites for that value. Most
low-frequency \evals are invoked only once; about half as many are invoked
twice; after that, the frequency quickly drops.

\Eval takes any value, but if its argument is not an expression, \eval returns
it unchanged. Expressions account for \packageCodepercent of arguments in our
corpus. More specifically, \packageSymbolpercent are \texttt{symbol}s (single
variables such as \c{x}), \packageLanguagepercent are \c{language} objects
(function calls such as \c{f(x)} or \c{x+1}), and \packageExpressionpercent are
\c{expression} objects (lists of expressions such as function calls or
symbols). Further inspection reveals that most symbols,
\packageGgplotsymbolpercent to be exact, come from a single site in the
\c{ggplot2} package and have the value \c{\_inherit}.\footnote{This models
inheritance in \c{ggproto}, one of the many object-oriented systems in
R, used exclusively in the \c{ggplot2} graphics library.}

To estimate how much executable code is injected through \eval, we measure the number of nodes in the expressions; for example, \c{x+1} counts as 3. We
also measured string lengths of unparsed expressions, but these measurements
were dominated by the size of data objects which could range in the MBs. The
median argument size is \packageMedianszeval (due to the symbols), and the
average is \packageAvgszeval nodes. The largest \eval input observed is
\packageMaxszeval\,-- a significant chunk of code.
Figure~\ref{fig:sizedistribution} shows the distribution of sizes for arguments
of fewer or equal to 25 nodes. The x-axis is the size of arguments in number of
nodes, and the y-axis is the count of arguments with that size. The size drops
rapidly, with few observations larger than 15 nodes. The long tail is omitted
for legibility.

To estimate the work performed in \evals, we count instructions executed by the
interpreter. Most invocations perform relatively little work, with
\packageSmalleventspct of \evals executing 50 or fewer instructions. The violin
plot of Figure~\ref{ev}(a) corresponds to \evals executing $\leq$ 50
instructions; it is dominated by trivial symbol lookups. Figure~\ref{ev}(b) has
the work-intensive \evals, which go all the way to \packageMaxeventsRnd
instructions.

@@

We recorded \baseAllcalls \eval calls in the \BasePackages base libraries. This
comes from running 10\% randomly selected programs from the
\CranRunnableScriptsRnd extracted programs from CRAN packages, covering
\baseTriggeredevalpct of the \BaseEvalCallSites sites. Most of the calls
(\baseEvalsratio) are to the \c{eval} function, and most arguments
(\baseCodepercent) are expressions. However, unlike CRAN, the majority
(\baseLanguagepercent) of the arguments are language objects. The median
argument size is \baseMedianszeval, which is more than for CRAN and the maximum size is \baseMaxszeval, less than for CRAN. Small instruction counts ($<50$) amount to \baseSmalleventspct of calls. A single site in the \c{match.arg}
function is responsible for \baseTopFuncPercent of all recorded calls. This
function provides a convenient way for argument verification using partial
matching. It is heavily relied on in R functions that use string arguments to
parameterize function behavior, a popular API mechanism in R. For example, in a
body of \c{center <- function(x, type = c("mean", "median", "trimmed"))} one
can get the value of \c{type} parameter using \c{type <- match.arg(type)}. A
call \c{center(x, "trim")} will match \c{type} to \c{"trimmed"}.

The challenge dealing with Base is that every program uses it. It can thus
generate extreme amounts of data, but the data is quite predictable. For the
rest of the paper, we only mention Base when there are surprising
observations.

@@

In total, \kaggleAllcalls \eval were recorded, all to \eval function. Out of
\kaggleStaticeval sites from \KaggleWithEvals, only \kaggleTriggeredeval sites
in \kaggleNbruns scripts were hit; \KaggleFailedScripts scripts failed to run,
and the other did not exercise the \eval sites. This is partially expected
as Kaggle code does not need to abide by any checks. Upon a manual inspection,
we observed that indeed the failing scripts were of poor quality, often not
finished, using misspelled package names, hard-coded file paths, or accessing
missing files.

Most call sites are invoked only once. Only one site is called \kaggleMaxcalls times.
Expressions amount for \kaggleCodepercent of arguments. Unlike the CRAN corpus,
there are very few symbols (\kaggleSymbolpercent). Most expressions result from
calls to \c{parse}; thus, most \evals start with strings. The median argument
size is \kaggleMedianszeval, which makes sense, as few arguments are only
symbols. The largest argument is \kaggleMaxszeval. The distribution of
instructions per eval is similar to CRAN; \kaggleSmalleventspct of \evals
execute fewer than 50 instructions. Manual inspection of \eval usage in Kaggle
suggests that it is consistent with the data obtained for CRAN. We do not
discuss it further.

@@

\subsection{Discussion}
These results show that \eval is central to the language implementation as it is
omnipresent in the relatively small Base library. The Base library is, in turn,
used by every single package. To avoid having Base dominate the result, we have
excluded calls occurring there from the data we report when analyzing packages.
CRAN Packages, which are typically developed by experienced programmers, make
regular and varied use of \eval. They represent our most interesting data set as
these packages result from over 20 years of contributions by thousands of
authors. The code in packages is well maintained and relatively well tested.
Finally, Kaggle scripts are often written by less sophisticated users and
likely perform more straightforward tasks, thus having a lesser need for \eval, and its uses originate from strings as these are easier to manipulate for end-users. Due to the relative paucity of data in Kaggle, we do not pursue the data set further.
